---
source: Rmd
title: "Using the dplyr package"
teaching: 0
exercises: 0
questions:
- "How do I select, filter, and subset data in R?"
objectives:
- "Select specific observations from data."
- "Filter out those observations that meet specific criteria."
- "Subset data."
- "Save transformed data."
keypoints:
- ""
- ""
---

```{r, results='hide'}
library(tidyverse)
```

## Reading text files (CSV)

Now that we know how to find our files, let's load up our first data set.
When trying to type in a file path,
you can hit the `<TAB>` key to autocomplete the files.
This will help you with a lot of potential spelling mistakes.

```{r}
read_csv("../data/medicaldata_tumorgrowth.csv")
```

**Debug help:**
- If the above code returns a `could not find function "read_csv"`
  make sure you have loaded up the proper library with `library(tidyverse)`
- If the above code returns a `does not exist in current working directory`,
  make sure the working directory it lists is your expected "starting point" (i.e., working directory),
  and make sure the file path is spelled correctly.

`read_csv` will show us the columns that were read in,
as well as the **data type** of that column (e.g., character, double -- a number).

Loading a data set is great, but we need a convenient way to refer to the data set.
We don't want to re-load the data set every time we want to perform an action on it.
We can take this loaded data set and **assign** it to a **variable**.
We can do this with the **assignment** **operator**, `<-`.
Note the way it is typed, a less than symbol (`<`) followed immediately by the dash (`-`) without any spaces in between.
The right side of the assignment operator, `<-`, will be executed and then **assigned** to the variable on the left.

```{r}
tumor <- read_csv("../data/medicaldata_tumorgrowth.csv")
```

Notice this time we no longer see the dataset being printed.
The "Environment" tab in the RStudio panel will now have an entry for the variable you used.
Clicking on the right data set icon will open a view of your dataset,
clicking on the arrow will show you the column-by-column text representation (technically it's called the `structure`).

To look at our dataset we can execute *just* the variable we assigned the dataset to.

```{r}
tumor
```

This tabular dataset that has now been loaded into R is called a **data frame** **object** (or simply **dataframe**),
the `tidyverse` uses a `tibble`.
For the most part, a `data.frame` object will behave like a `tibble` object.


## What is dplyr?
The package dplyr provides easy tools for the most common data manipulation tasks. It is built to work directly 
with data frames, with many common tasks optimized by being written in a compiled language (C++). An additional 
feature is the ability to work directly with data stored in an external database. The benefits of doing this 
are that the data can be managed natively in a relational database, queries can be conducted on that database, 
and only the results of the query are returned.

This addresses a common problem with R in that all operations are conducted in-memory and thus the amount of 
data you can work with is limited by available memory. The database connections essentially remove that 
limitation in that you can connect to a database of many hundreds of GB, conduct queries on it directly, 
and pull back into R only what you need for analysis.

To learn more about dplyr after the workshop, you may want to check out this handy 
[data transformation with dplyr cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-transformation.pdf).

## Selecting columns

Here is the tumor dataset we are working with.

```{r}
tumor
```

In order to subset the columns of our dataset, we can use the `select` function.
The **first** **argument** of the `select` function is the dataset we want to select columns from.
Then, the other arguments are the columns we want to select.


```{r}
select(tumor, Size)
```

If we want multiple columns (or to re-order columns) we can specify each column we want selected.

```{r}
select(tumor, Group, Day, Size)
```

### The `%>%` pipe operator

A common way `tidyverse` code gets written is using the `%>%` operator.
The `%>%` takes the object on the left, and passes that into the first argument to the function on the right.
We can re-write the above select statements as follows.
R interprets the code in the exact same way.

```{r}
tumor %>%
  select(Size)
```

The way we pronounce the pipe, `%>%` is "then".
So the above statement is read "tumor, then, select size".

```{r}
tumor %>%
  select(Group, Day, Size)
```

The benefits of the pipe may not be as apparent now,
but when we start chaining multiple data sets together,
it'll make the code a little more readable.

## Filtering rows

Now that we know how to select columns on our dataframe,
we can now learn how to filter the rows.

To filter rows, we can use the `filter` function.
The first argument of `filter` is the dataframe object (the same pattern as the `select` function),
and the other arguments are statements that return some `TRUE` or `FALSE` value (this is known as a **boolean** condition).

For example if we want to filter the `tumor` dataset such that the `Group` value is `1` we can filter the rows using that condition.

```{r}
filter(tumor, Group == 1)
```

We can write the "pipe" version as such:

```{r}
tumor %>%
  filter(Group == 1)
```

If we want to combine multiple conditions, where `Group` is `1` **and** `Day` is `0`,
we can put each statement as a separate parameter into the `filter` function.

```{r}
tumor %>%
  filter(Group == 1, Day == 0)
```

We can also use the `&` for the **and** operator.

```{r}
tumor %>%
  filter(Group == 1 & Day == 0)
```

Or use the `|` for the **or** operator.

```{r}
tumor %>%
  filter(Day == 0 | Day == 13)
```


## Subsetting columns and rows

To subset on both columns **and** rows,
we call the corresponding subseting function one after the other.
Here we are taking the `tumor` column, then `select`ing the `Group`, `Day` and `Size` columns, then
`filter`ing the rows where the `Size` is greater than `2000`.

```{r}
tumor %>%
  select(Group, Day, Size) %>%
  filter(Size > 2000)
```

This is an example where the pipe, `%>%` makes the code a little easier to **read**.
We can "un-pipe" our code and  it looks like this:

```{r}
filter(select(tumor, Group, Day, Size), Size > 2000)
```


## Saving out data

Saving out your data frame objects is the first step in creating pipelines.
Even if all you do is read in data, filter it, and write out the subsetted dataset for another use case,
you have created your first pipeline.
Now, you have a repeatable way of performing the same action when your input dataset is updated or changed.

Let's say we want to work with a filtered version of our `tumor` dataset,
we can first write the code to subset the data we want.

```{r}
filtered <- tumor %>%
  select(Group, ID, Day, Size) %>%
  filter(Day == 0 | Day == 13)
```

```{r}
filtered
```

We can then write out the datafrme to a file (in this case a CSV text file),
using the same file pathing string format as we loaded a dataset earlier.

```{r}
write_csv(filtered, "data/tumor_filtered.csv")
```

One thing to be mindful of is most programs and computer languages don't like having spaces in file names.
In general, you'll end up typing more characters dealing with spaces in file names,
instead of using an underscore, `_`, or dash, `-`.
Also as a general rule, keeping files in lower-case will prevent a lot of potential spelling mistakes.

<button class="accordion" id="question">Exercise 3 Question</button>
:::{.panel-question}
1. Read in all the patient data from the spreadsheets chapter (chapter \@ref(spreadsheets)) found in `data > spreadsheets > all.csv`
2. `select` the `first_name`, `last_name`, `day`, `time`, and `temp_f` columns.
3. `filter` the rows such that the person's name is "Aniya Warner"
4. Save the filtered dataset into the `output` folder named `aniya_warner.csv`.
:::

## Summary {#load-data-summary}

## Additional Resources {#load-data-resources}
