---
source: Rmd
title: "Data Transformations"
teaching: 0
exercises: 0
questions:
- "How do I obtain transform data?"
objectives:
- "Add new columns to a data frame that are functions of existing columns with `mutate`."
- "Use the split-apply-combine concept for data analysis."
- "Use `summarize`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results."
keypoints:
- ""
- ""
---

### Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column of tumor size (volume) in cubic centimeters:

```{r, purl = FALSE}
tumor %>%
  mutate(size_cubic_cm = Size / 1000)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl = FALSE}
tumor %>%
  mutate(size_cubic_cm = Size / 1000,
         size_cubic_in = size_cubic_cm / 16)
```

If this runs off your screen and you just want to see the first few rows, you
can use a pipe to view the `head()` of the data. (Pipes work with non-**`dplyr`**
functions, too, as long as the **`dplyr`** or `magrittr` package is loaded).

```{r, purl = FALSE}
tumor %>%
  mutate(size_cubic_cm = Size / 1000) %>%
  head()
```

> ### Challenge {.challenge}
>
>  Create a new data frame from the `surveys` data that meets the following
>  criteria: contains only the `Grp` column and a new column called
>  `size_cubic_in` containing the `Size` values converted to cubic inches.
>  In this `size_cubic_in` column, all values are greater than 0.1.
>
>  **Hint**: think about how the commands should be ordered to produce this data frame!
> 
> ```{r, answer=TRUE, eval=FALSE, purl=FALSE}
> tumor_cubic_in <- tumor %>%
>     mutate(size_cubic_cm = Size / 1000,
         size_cubic_in = size_cubic_cm / 16) %>%
>     filter(size_cubic_in > 0.1) %>%
>     select(Grp, size_cubic_in)
> ```

### Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of the
`group_by()` function.


#### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the mean `weight` by sex:

```{r, purl = FALSE}
tumor %>%
  group_by(Grp) %>%
  summarize(mean_size = mean(Size))
```

You can also group by multiple columns:

```{r, purl = FALSE}
tumor %>%
  group_by(Grp, Day) %>%
  summarize(mean_size = mean(Size)) %>% 
  tail()
```

Here, we used `tail()` to look at the last six rows of our summary. Before, we had 
used `head()` to look at the first six rows. 

If you want to display more data, you can use the `print()` function
at the end of your chain with the argument `n` specifying the number of rows to
display:

```{r, purl = FALSE}
tumor %>%
  group_by(Grp, Day) %>%
  summarize(mean_size = mean(Size)) %>% 
  print(n = 15)
```

Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum weight for each species for each sex:

```{r, purl = FALSE}
tumor %>%
  group_by(Grp, Day) %>%
  summarize(mean_size = mean(Size),
            min_size = min(Size))
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of mean weight:

```{r, purl = FALSE}
tumor %>%
  group_by(Grp, Day) %>%
  summarize(mean_size = mean(Size),
            min_size = min(Size)) %>%
  arrange(desc(mean_size))
```


#### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each group, we would do:

```{r, purl = FALSE}
tumor %>%
    count(Grp) 
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `tumor %>% count()` is equivalent to:  

```{r, purl = FALSE}
tumor %>%
    group_by(Grp) %>%
    summarise(count = n())
```

For convenience, `count()` provides the `sort` argument:  

```{r, purl = FALSE}
tumor %>%
    count(Grp, sort = TRUE) 
```

Previous example shows the use of `count()` to count the number of rows/observations 
for *one* factor (i.e., `Grp`). 
If we wanted to count *combination of factors*, such as `Grp` and `Day`, 
we would specify the first and the second factor as the arguments of `count()`:

```{r purl = FALSE}
tumor %>%
  count(Grp, Day) 
```

With the above code, we can proceed with `arrange()` to sort the table 
according to a number of criteria so that we have a better comparison. 
For instance, we might want to arrange the table above in (i) an alphabetical order of 
the levels of the group and (ii) in descending order of the count:

```{r purl = FALSE}
tumor %>%
  count(Grp, Day) %>%
  arrange(Day, desc(n))
```

> ### Challenge {.challenge}
>
> 1. How many observations are there for each `ID` number?
>
> ```{r, answer=TRUE, purl=FALSE}
> tumor %>%
>     count(ID) 
> ```
>
> 2. Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
> length for each species (using `species_id`). Also add the number of
> observations (hint: see `?n`).
>
> ```{r, answer=TRUE, purl=FALSE}
> tumor %>%
>     filter(!is.na(hindfoot_length)) %>%
>     group_by(species_id) %>%
>     summarize(
>         mean_hindfoot_length = mean(hindfoot_length),
>         min_hindfoot_length = min(hindfoot_length),
>         max_hindfoot_length = max(hindfoot_length),
>         n = n()
>     )
> ```
>
> 3. What was the heaviest animal measured in each year? Return the columns `year`,
> `genus`, `species_id`, and `weight`.
>
> ```{r, answer=TRUE, purl=FALSE}
> tumor %>%
>     filter(!is.na(weight)) %>%
>     group_by(year) %>%
>     filter(weight == max(weight)) %>%
>     select(year, genus, species, weight) %>%
>     arrange(year)
> ```

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Count Challenges:
##  1. How many animals were caught in each `plot_type` surveyed?
##  2. Use `group_by()` and `summarize()` to find the mean, min, and max
## hindfoot length for each species (using `species_id`). Also add the number of
## observations (hint: see `?n`).
##  3. What was the heaviest animal measured in each year? Return the
##  columns `year`, `genus`, `species_id`, and `weight`.
```
